---
title: "STAT545 hw03"
output: github_document
---

<style type="text/css">
.twoC {width: 100%}
.twoCl {width: 100%}
.clearer {clear: both}
.twoC table {max-width: 35%; float: left; max-height: 430px}
.twoC img {max-width: 65%; float: right}
.twoCl table {max-width: 35%; float: left; max-height: 860px}
.twoCl img {max-width: 65%; float: right}
</style>

## Setting ups

First we need to install and load packages.
```{r}
#install.packages("tidyverse")
#install.packages("gapminder")
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
```

Now we are ready to go!

## Tasks

### Get the maximum and minimum of GDP per capita for all continents.

```{r eval=FALSE}
a <- gapminder %>%
  group_by(continent) %>%
  summarise(min = min(gdpPercap),
            max = max(gdpPercap)) %>%
  gather("type", "value", min:max)

knitr::kable(a)

a %>% ggplot(aes(x = continent, fill = type)) + 
  geom_bar(aes(y = value), stat = "identity", position="dodge") +
  scale_y_log10()
```

<div class="twoC">
```{r results = 'asis', echo=FALSE}
a <- gapminder %>%
  group_by(continent) %>%
  summarise(min = min(gdpPercap),
            max = max(gdpPercap)) %>%
  gather("type", "value", min:max)

knitr::kable(a)

a %>% ggplot(aes(x = continent, fill = type)) + 
  geom_bar(aes(y = value), stat = "identity", position="dodge") +
  scale_y_log10()
```
</div><div class="clearer"></div>


### Look at the spread of GDP per capita within the continents.

```{r eval=FALSE}
a <- gapminder %>%
  group_by(continent) %>%
  select(continent, gdpPercap) %>%
  summarise(min = min(gdpPercap),
            max = max(gdpPercap),
            mean = mean(gdpPercap),
            Qu1 = quantile(gdpPercap, 0.25),
            median = median(gdpPercap),
            Qu3 = quantile(gdpPercap, 0.75))

knitr::kable(a)

gapminder %>%
  ggplot(aes(x = gdpPercap, group = continent, color = continent)) +
  geom_density() +
  scale_x_log10()
```

<div class="twoC">
```{r results = 'asis', echo=FALSE}
a <- gapminder %>%
  group_by(continent) %>%
  select(continent, gdpPercap) %>%
  summarise(min = min(gdpPercap),
            max = max(gdpPercap),
            mean = mean(gdpPercap),
            Qu1 = quantile(gdpPercap, 0.25),
            median = median(gdpPercap),
            Qu3 = quantile(gdpPercap, 0.75))

knitr::kable(a)

gapminder %>%
  ggplot(aes(y = gdpPercap, x = continent)) +
  geom_violin() +
  scale_y_log10() +
  geom_jitter(aes(size = pop), alpha = 0.2)
```
</div><div class="clearer"></div>

### Compute a mean life expectency weighted by population. 

```{r eval=FALSE}
a <- gapminder %>%
  group_by(year) %>%
  summarise(wtd_lifeExp = weighted.mean(lifeExp, pop))

knitr::kable(a)

a %>% ggplot(aes(year, wtd_lifeExp)) + 
  geom_point() + 
  geom_line()
```

<div class="twoC">
```{r results = 'asis', echo=FALSE}
a <- gapminder %>%
  group_by(year) %>%
  summarise(wtd_lifeExp = weighted.mean(lifeExp, pop))

knitr::kable(a)

a %>% ggplot(aes(year, wtd_lifeExp)) + 
  geom_point() + 
  geom_line()
```
</div><div class="clearer"></div>

### How is life expectancy changing over time on different continents?

```{r eval=FALSE}
a <- gapminder %>%
  group_by(continent, year) %>%
  summarise(wtd_lifeExp = weighted.mean(lifeExp, pop))

knitr::kable(a)

a %>% ggplot(aes(year, wtd_lifeExp)) + 
  geom_point(aes(color = continent)) + 
  geom_line(aes(color = continent)) +
  geom_smooth(method = "loess")

```

<div class="twoC">
```{r results = 'asis', echo=FALSE}
a <- gapminder %>%
  group_by(continent, year) %>%
  summarise(wtd_lifeExp = weighted.mean(lifeExp, pop))

knitr::kable(a)

a %>% ggplot(aes(year, wtd_lifeExp)) + 
  geom_point(aes(color = continent)) + 
  geom_line(aes(color = continent)) +
  geom_smooth(method = "loess")

```
</div><div class="clearer"></div>

### Report the absolute and/or relative abundance of countries with low life expectancy over time by continent.

```{r eval=FALSE}
a <- gapminder %>%
  group_by(continent, year) %>%
  mutate(wtd_lifeExp = weighted.mean(lifeExp, pop)) %>%
  summarise(low_ct = sum(lifeExp < wtd_lifeExp),
            low_ct_rl = low_ct/length(country))

knitr::kable(a)

a %>% ggplot(aes(year, low_ct)) + 
  geom_point(aes(color = continent)) + 
  geom_line(aes(color = continent))

a %>% ggplot(aes(year, low_ct_rl)) + 
  geom_point(aes(color = continent)) + 
  geom_line(aes(color = continent))
```
<div class="twoCl">
```{r results = 'asis', echo=FALSE}
a <- gapminder %>%
  group_by(continent, year) %>%
  mutate(wtd_lifeExp = weighted.mean(lifeExp, pop)) %>%
  summarise(low_ct = sum(lifeExp < wtd_lifeExp),
            low_ct_rl = low_ct/length(country))

knitr::kable(a)

a %>% ggplot(aes(year, low_ct)) + 
  geom_point(aes(color = continent)) + 
  geom_line(aes(color = continent))

a %>% ggplot(aes(year, low_ct_rl)) + 
  geom_point(aes(color = continent)) + 
  geom_line(aes(color = continent))
```
</div><div class="clearer"></div>

### Find countries with interesting stories.

Interesting stories always come from changes. So first I calculate the range of life expectency, population, gdp per capita and whole gdp for each country.
```{r}
a <- gapminder %>%
  mutate(gdp = gdpPercap * pop) %>%
  group_by(country) %>%
  summarise(range_life = max(lifeExp) - min(lifeExp),
            range_pop = max(pop) - min(pop),
            range_gpc = max(gdpPercap) - min(gdpPercap),
            range_gdp = max(gdp) - min(gdp))
```

Now we can check the data and plot for the most various countries.
```{r, eval=FALSE}
knitr::kable(head(arrange(a, desc(range_life))))
range_lif <- filter(gapminder, country %in% select(head(arrange(a, desc(range_life))), country)$country)
range_lif %>% ggplot(aes(year, lifeExp, color=country)) + geom_point() + geom_line()
```
```{r, results='asis', echo=FALSE}
knitr::kable(head(arrange(a, desc(range_life))))
range_lif <- filter(gapminder, country %in% select(head(arrange(a, desc(range_life))), country)$country)
range_lif %>% ggplot(aes(year, lifeExp, color=country)) + geom_point() + geom_line()
```

```{r, eval=FALSE}
knitr::kable(head(arrange(a, desc(range_pop))))
range_pop <- filter(gapminder, country %in% select(head(arrange(a, desc(range_pop))), country)$country)
range_pop %>% ggplot(aes(year, pop, color=country)) + geom_point() + geom_line()
```
```{r, results='asis', echo=FALSE}
knitr::kable(head(arrange(a, desc(range_pop))))
range_pop <- filter(gapminder, country %in% select(head(arrange(a, desc(range_pop))), country)$country)
range_pop %>% ggplot(aes(year, pop, color=country)) + geom_point() + geom_line()
```

```{r, eval=FALSE}
knitr::kable(head(arrange(a, desc(range_gpc))))
range_gpc <- filter(gapminder, country %in% select(head(arrange(a, desc(range_gpc))), country)$country)
range_gpc %>% ggplot(aes(year, gdpPercap, color=country)) + geom_point() + geom_line()
```
```{r, results='asis', echo=FALSE}
knitr::kable(head(arrange(a, desc(range_gpc))))
range_gpc <- filter(gapminder, country %in% select(head(arrange(a, desc(range_gpc))), country)$country)
range_gpc %>% ggplot(aes(year, gdpPercap, color=country)) + geom_point() + geom_line()
```

```{r, eval=FALSE}
knitr::kable(head(arrange(a, desc(range_gdp))))
range_gdp <- filter(gapminder, country %in% select(head(arrange(a, desc(range_gdp))), country)$country)
range_gdp %>% ggplot(aes(year, gdpPercap * pop, color=country)) + geom_point() + geom_line()
```
```{r, results='asis', echo=FALSE}
knitr::kable(head(arrange(a, desc(range_gdp))))
range_gdp <- filter(gapminder, country %in% select(head(arrange(a, desc(range_gdp))), country)$country)
range_gdp %>% ggplot(aes(year, gdpPercap * pop, color=country)) + geom_point() + geom_line()
```

